\documentclass[11pt]{article}

% Switch ``review'' to ``final'' for the camera-ready version.
\usepackage[review]{acl}

\usepackage{times}
\usepackage{latexsym}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{microtype}
\usepackage{inconsolata}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{multirow}

% TODO: replace [Team Name] before camera-ready.
\title{vl4dio4n at \#SMM4H-HeaRD 2026 Task~7:\\
  Comparing Encoder-, LLM-, and SVM-Based NER Pipelines for
  Substance-Use Impact Detection}

\author{Anonymous Submission}

\begin{document}
\maketitle

\begin{abstract}
We describe our submission to SMM4H-HeaRD~2026 Task~7, which asks
systems to label \texttt{ClinicalImpacts} and \texttt{SocialImpacts}
spans in Reddit posts about non-medical substance use. We compare
four pipeline shapes built on the same DeBERTa-v3-base backbone:
(i)~a direct $5$-class encoder with a linear-chain CRF head, (ii)~a
two-stage detect-then-classify pipeline that delegates span typing
to an instruction-tuned LLM (Qwen2.5-7B or Gemma-3-12B, $4$-bit
NF4), (iii)~an audit pipeline in which the same LLM verifies the
encoder's predictions, and (iv)~a classical-ML variant that replaces
the LLM with an SVM trained on encoder span embeddings. Across
$16$~configurations, the encoder-only \emph{DeBERTa-v3 + CRF}
configuration is the strongest single system on the official test
split, reaching $45.4\%$ strict and $54.2\%$ relaxed F1 --- $+8.6$ /
$+5.3$ points above a \texttt{mental-roberta-base} baseline. LLM
audits give a small dev gain that does not transfer to test.
\end{abstract}

\section{Introduction}
\label{sec:introduction}

SMM4H-HeaRD~2026 Task~7~\cite{dey2025inference} releases a corpus of
Reddit posts about non-medical opioid use, annotated under a
$5$-label BIO scheme that distinguishes
\texttt{ClinicalImpacts} (e.g.\ \emph{withdrawal}, \emph{rehab})
from \texttt{SocialImpacts} (e.g.\ \emph{lost my job},
\emph{homeless}). Systems are scored with strict micro-F1 and a
relaxed overlap F1 in the CoNLL-2003
style~\cite{tjong2003conll}, computed by \texttt{seqeval}.

Our submission is a comparative study, not a single architectural
proposal. We fix the encoder backbone to \texttt{DeBERTa-v3-base}
\cite{he2023debertav3} and vary what sits on top of it, asking two
questions. \textbf{RQ1.} Does decoupling span detection from span
typing improve over a well-decoded $5$-class CRF? \textbf{RQ2.} Does
adding an LLM or SVM auditor on top of the CRF recover any
appreciable F1?

We answer both questions with a $16$-row ablation
(Table~\ref{tab:ablation}). The encoder-only DeBERTa-v3 + CRF
configuration is the strongest \emph{single} system on the test
split, and none of the LLM- or SVM-augmented variants improves on
it. A $4$-shot Gemma-3-12B audit obtains the best dev strict F1 but
loses to the encoder on test, which we attribute to the
distributional shift between dev and test on this corpus.
% The full source code, configuration files and trained-model
% checkpoints are available at
% \url{https://github.com/vl4dio4n/SMM4H-HeaRD-2026-Task-7}
% (URL anonymised during review; uncomment for the camera-ready version).

\section{Data}
\label{sec:data}

The released corpus consists of $842$ training posts ($\sim$$17.2$K
tokens), $258$ development posts ($\sim$$5.2$K tokens), and $278$
test posts ($\sim$$6.2$K tokens). The training and development CSVs
contain (mostly) one row per post; the test CSV is aggressively
segmented into $1.7$K sentence-level rows, which forces inference to
run on shorter contexts than training. The $5$ BIO labels are
heavily imbalanced: the \texttt{O} tag accounts for $\sim$$94\%$ of
all training tokens (a $\sim$$16{:}1$ ratio against entity tokens).
On train and dev, \texttt{ClinicalImpacts} spans outnumber
\texttt{SocialImpacts} spans by $\sim$$3{:}1$, but on the test set
the ratio inverts and \texttt{SocialImpacts} becomes the majority
class --- a shift we revisit in
Sec.~\ref{sec:results}. The test set is also noticeably more
first-person ($\sim$$71\%$ of segments) than train or
dev ($\sim$$45\%$).

\section{System Description}
\label{sec:system}

We compare four pipelines, all built around the same DeBERTa-v3
backbone fine-tuned with a maximum sequence length of $256$, batch
size $16$, $20$ epochs, $10\%$ linear warm-up and a base learning
rate of $2{\times}10^{-5}$.

\paragraph{Direct + CRF (Exp~04, our main system).} The encoder
predicts $5$-class BIO tags directly. A linear-chain
CRF~\cite{lafferty2001crf,lample2016neural} on top of the encoder
replaces token-level cross-entropy with the CRF negative
log-likelihood and decodes via Viterbi at inference. The CRF rules
out illegal BIO transitions, which mainly cleans up boundaries (and
hence strict F1) without changing the set of detected entity
regions. We add Layer-wise Learning-Rate
Decay~\cite{howard2018ulmfit} with multiplier $0.9$ to stabilise
DeBERTa fine-tuning. This configuration is our final Codabench
submission.

\paragraph{Two-stage with LLM span typer (Exp~05a, 07a, 08a, 09a).}
The encoder is re-trained on a collapsed $3$-class label set
($\{\texttt{O}, \texttt{B-Imp}, \texttt{I-Imp}\}$) and only detects
span boundaries; an instruction-tuned LLM types each predicted span
as \texttt{ClinicalImpacts}, \texttt{SocialImpacts}, or
\texttt{Neither}. We use focal loss~\cite{lin2017focal} ($\gamma{=}2$)
to handle class imbalance in the detector, since the CRF is no
longer in charge of the $5$-class structure.

\paragraph{Audit with LLM verifier (Exp~05b, 07b, 08b, 09b).} The
$5$-class DeBERTa-v3 + CRF detector of Exp~04 produces a full BIO
tagging; the LLM is shown the encoder's predicted span and label as
a hint and may verify, correct (Clinical~$\leftrightarrow$~Social),
or reject it (mark as Neither, which removes the span). Audit and
two-stage share the same prompt template and class
descriptions; only the candidate construction differs.

\paragraph{LLM details.} We use
\texttt{Qwen/Qwen2.5-7B-Instruct} and
\texttt{google/gemma-3-12b-it}, loaded in $4$-bit NF4 precision via
\texttt{bitsandbytes}~\cite{dettmers2023qlora} so that both fit on a
single consumer GPU. Decoding is greedy and capped at $6$ new
tokens; the post is truncated to $2{,}000$ characters and the
candidate span to $200$. We compare zero-shot prompting with
$4$-shot in-context learning~\cite{brown2020fewshot}; few-shot
demonstrations are mined from gold training spans and balanced
across the three classes (one per class plus one randomly drawn
\texttt{Neither}).

\paragraph{SVM-augmented variant (Exp~10a/10b).} To isolate how much
of the LLM's contribution comes from world knowledge that is not
already linearly accessible from the encoder, we replace the LLM
with an SVM~\cite{cortes1995svm}. Each candidate span is represented
by the mean-pooled hidden state of its tokens in the (frozen)
fine-tuned encoder; a
\texttt{StandardScaler}\,$\rightarrow$\,\texttt{PCA(128)}\,$\rightarrow$\,\texttt{SVC(RBF, class\_weight=balanced)}
pipeline is fitted with $5$-fold grid-searched macro-F1 over $C$
and $\gamma$.

\paragraph{Baseline.} For comparability we keep the
\texttt{mental-roberta-base}~\cite{ji2022mentalbert} +
cross-entropy, $5$-class direct configuration as our internal
baseline (Exp~01); it is a RoBERTa~\cite{liu2019roberta} variant
adapted to mental-health Reddit text and is the natural starting
point given the domain.

\section{Experiments}
\label{sec:experiments}

We report the full $16$-row ablation in
Table~\ref{tab:ablation}. Each row pairs a pipeline shape (Pipe:
\texttt{D}~direct, \texttt{2S}~two-stage, \texttt{Au}~audit,
\texttt{Ens}~ensemble, \texttt{2S/Au+SVM}~SVM-augmented
two-stage/audit), a backbone, a loss function, and an external span
classifier when applicable. ``Dev'' columns are computed on the
official $258$-row development split with \texttt{seqeval};
``Test'' columns are the official Codabench leaderboard scores
returned for the corresponding submission.

\begin{table*}[!t]
\centering
\footnotesize
\setlength{\tabcolsep}{3.5pt}
\renewcommand{\arraystretch}{1.05}
\begin{tabular}{l l l l c l c c c c}
\toprule
\textbf{Exp} & \textbf{Pipe} & \textbf{Backbone} & \textbf{Loss}
& \textbf{CRF} & \textbf{External classifier}
& \textbf{Dev~S} & \textbf{Dev~R} & \textbf{Test~S} & \textbf{Test~R} \\
\midrule
01  & D    & mental-roberta-base        & CE      & N & ---                       & 0.356          & 0.489          & 0.368          & 0.489 \\
02a & D    & mental-roberta-base        & W-CE    & N & ---                       & 0.360          & 0.493          & 0.410          & 0.549 \\
02b & D    & mental-roberta-base        & Focal   & N & ---                       & 0.339          & 0.492          & 0.336          & 0.507 \\
03  & D    & deberta-v3-base            & Focal   & N & ---                       & 0.345          & 0.539          & 0.424          & \textbf{0.558} \\
04  & D    & deberta-v3-base            & CRF NLL & Y & ---                       & 0.413          & \textbf{0.571} & \textbf{0.454} & 0.542 \\
06  & Ens  & deberta-v3-base $\times 3$ & CRF NLL & Y & ---                       & 0.410          & 0.565          & 0.441          & 0.551 \\
\midrule
05a & 2S   & deberta-v3-base (3-cls)    & Focal   & N & Qwen2.5-7B, zero-shot     & 0.394          & 0.423          & 0.421          & 0.449 \\
05b & Au   & deberta-v3-base            & CRF NLL & Y & Qwen2.5-7B, zero-shot     & 0.386          & 0.541          & 0.443          & 0.549 \\
07a & 2S   & deberta-v3-base (3-cls)    & Focal   & N & Qwen2.5-7B, $4$-shot      & 0.381          & 0.431          & 0.434          & 0.483 \\
07b & Au   & deberta-v3-base            & CRF NLL & Y & Qwen2.5-7B, $4$-shot      & 0.380          & 0.537          & 0.440          & 0.552 \\
08a & 2S   & deberta-v3-base (3-cls)    & Focal   & N & Gemma-3-12B, zero-shot    & 0.382          & 0.421          & 0.395          & 0.418 \\
08b & Au   & deberta-v3-base            & CRF NLL & Y & Gemma-3-12B, zero-shot    & 0.388          & 0.550          & 0.400          & 0.496 \\
09a & 2S   & deberta-v3-base (3-cls)    & Focal   & N & Gemma-3-12B, $4$-shot     & 0.408          & 0.438          & 0.428          & 0.444 \\
09b & Au   & deberta-v3-base            & CRF NLL & Y & Gemma-3-12B, $4$-shot     & \textbf{0.426} & 0.567          & 0.432          & 0.523 \\
\midrule
10a & 2S+SVM & deberta-v3-base (3-cls)  & Focal   & N & SVM (RBF, PCA-128)        & 0.357          & 0.367          & 0.433          & 0.435 \\
10b & Au+SVM & deberta-v3-base          & CRF NLL & Y & SVM (RBF, PCA-128)        & 0.407          & 0.559          & ---            & ---   \\
\bottomrule
\end{tabular}
\caption{Ablation across all $16$ configurations. Pipelines:
\texttt{D}~=~direct, \texttt{2S}~=~two-stage, \texttt{Au}~=~audit,
\texttt{Ens}~=~ensemble, \texttt{2S/Au+SVM}~=~SVM-augmented. ``Dev''
and ``Test'' columns give strict (S) and relaxed (R) F1 on the
development split and the Codabench test split, respectively.
Layer-wise LR Decay is enabled for all DeBERTa-v3 runs (Exp~03 and
above). Best per metric in \textbf{bold}.}
\label{tab:ablation}
\end{table*}

\section{Results and Discussion}
\label{sec:results}

\paragraph{Encoder choices and CRF dominate.} Switching the
backbone from \texttt{mental-roberta-base} (Exp~01) to
\texttt{deberta-v3-base} with focal loss and LLRD (Exp~03) is the
single largest step on test relaxed F1 ($+6.9$ over Exp~01).
Adding the CRF head (Exp~04) gives a further $+3.0$ test strict
points and reduces the strict-relaxed gap from $19.4$ pts (Exp~03,
dev) to $8.8$ pts on test, confirming that the CRF mainly cleans up
span boundaries rather than discovering new entities. Exp~04 is the
only configuration that exceeds $0.45$ test strict F1 across the
entire ablation.

\paragraph{LLM auditing helps on dev, not on test.} The strongest
LLM configuration is \emph{audit + Gemma-3-12B + 4-shot} (Exp~09b),
which reaches $0.426$ dev strict F1 --- the highest in the table ---
but only $0.432$ test strict, $\sim$$2$ points below Exp~04. The
two-stage shapes consistently lose to the audit shapes at relaxed
F1 because the $3$-class detector finds slightly worse spans than
the $5$-class CRF and the LLM cannot fix upstream boundary errors.
Within prompting regimes, $4$-shot in-context demonstrations help
Gemma (e.g.\ $+3.8$ dev strict on the audit pipeline) but slightly
hurt Qwen ($-0.6$ dev strict), suggesting that the larger model
extracts more signal from demonstrations.

\paragraph{SVM auditing is essentially neutral.} The audit SVM
(Exp~10b) preserves the Exp~04 dev numbers ($0.407$ strict /
$0.559$ relaxed) and reaches a $5$-fold macro-F1 of $1.0$ on gold
training embeddings: the two impact classes are linearly separable
in the CRF encoder's feature space, so the SVM rarely overrides
the encoder. This is the cleanest evidence in our ablation that
the LLM's added value over the CRF is bounded by the linear
information already present in the encoder representation.

\paragraph{Recall is the bottleneck.} Across the entire ablation,
relaxed precision climbs from $0.58$ (Exp~01) to $0.73$ (Exp~09b)
on dev, while relaxed recall stays in the $0.42$--$0.48$ band. Our
systems become progressively more conservative but do not discover
materially more gold spans. Combined with the train/dev~vs.~test
class flip (\texttt{ClinicalImpacts} majority on train/dev,
\texttt{SocialImpacts} majority on test), this explains why the
dev-best configuration (Exp~09b) is not the test-best
configuration (Exp~04).

\paragraph{Comparison to the organisers' baseline.} Our final
submission (Exp~04) is $\sim$$7$ relaxed-F1 points behind the
DeBERTa-large baseline of $\sim$$0.61$ relaxed F1 reported by
\citet{dey2025inference}. The most plausible drivers of the gap
are: (i)~we use the \emph{base} variant of every backbone we tried,
(ii)~our pipeline processes each released CSV row independently and
therefore cannot recover entities that span multiple sentence-level
test segments, and (iii)~our $4$-shot demonstrations are sampled by
class only, with no balancing for narrative perspective or span
length.

\section{Conclusion}
\label{sec:conclusion}

We presented a controlled comparison of four NER pipeline shapes on
SMM4H-HeaRD~2026 Task~7. The strongest single system on the
official test split is a $5$-class DeBERTa-v3 + CRF encoder
(Exp~04, $0.454$ strict / $0.542$ relaxed F1). LLM and SVM auditing
on top of the same encoder bring at most a small dev gain that does
not transfer to test, mostly because the test split shifts to a
more first-person, \texttt{Social}-heavy distribution that the
class-only $4$-shot sampler cannot anticipate. Recall on rare
\texttt{SocialImpacts} expressions remains the dominant failure
mode, suggesting that a larger backbone, span-merging across
sentence-level test segments, and perspective-aware in-context
retrieval are the most promising directions for future work.

\section*{Limitations}
We use only $4$-shot in-context demonstrations and do not fine-tune
the LLMs on the auditing task; parameter-efficient fine-tuning on
the training spans may close the gap to the organisers' baseline.
Our SVM head is restricted to linear-RBF kernels with
$128$ PCA components; richer span representations
(e.g.\ concatenation with first/last token embeddings) were not
explored. Each released CSV row is processed independently, so our
pipeline cannot recover impact spans that cross sentence-level
test segments.

\bibliography{custom}

\end{document}
